---
title: "Options Part 4"
author: "Pritam Dalal"
output: 
  beamer_presentation:
    theme: "default"
    colortheme: "fly"
    fonttheme: "default"
    df_print: tibble
header-includes:
   - \usepackage{amsmath}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(message = FALSE)
```

## Calculating PNL (1 of 8)

- Suppose you trade an option $\mathcal{O}$ at a price $P$ on trade-date $T_{1}$.
\vspace{10pt}
- Suppose you hold the option until expiration, which is trade-date $T_{n}$.
\vspace{10pt}
- The letter $i$ will serve as an index over the trade-dates, so $i = 1,\ldots,n$. 


## Calculating PNL (2 of 8)
- Let $B_{i}$ and $A_{i}$ be the end-of-day bid/ask prices of the option for trade-date $T_{i}$.  
\vspace{10pt}
- Note that $B_{n} = A_{n} = \text{option-payoff}$.
\vspace{10pt}
- $D_{i}$ - daily PNL for the trade as of end-of-day $T_{i}$.
\vspace{10pt}
- $C_{i}$ - trade-to-date (cummulative) PNL for the trade as of end-of-day $T_{i}$.
\vspace{10pt}
- **Intuition:** The cummulative PNL on a trade is how much money you make if you unwind the trade at current market values.


## Calculating PNL (3 of 8)

*Cummulative as Sum of Daily:* **BUY**


\begin{align*}
D_{i} &= \begin{cases} B_{1} - P & i = 1 \\ B_{i} - B_{i-1} & i > 1 \end{cases} \\[10pt]
C_{i} &= \sum_{k=1}^{i} D_{k}
\end{align*}

&nbsp;

**Exercise:** Show that $C_{j} = B_{j} - P$. 


## Calculating PNL (4 of 8)
 
*Cummulative as Sum of Daily:*  **SELL**


\begin{align*}
D_{i} &= \begin{cases} P - A_{1} & i = 1 \\ A_{i-1} - A_{i} & i > 1 \end{cases} \\[10pt]
C_{i} &= \sum_{k=1}^{i} D_{k}
\end{align*}

&nbsp;

**Exercise:** Show that $C_{i} = P - A_{i}$. 



## Calculating PNL (5 of 8)

*Daily as Change in Cummulative:* **BUY**


\begin{align*}
C_{i} &= B_{i} - P \\[10pt]
D_{i} &= \begin{cases} C_{1} & i = 1 \\ C_{i} - C_{i-1} & i > 1 \end{cases}
\end{align*}

&nbsp;

**Exercise:** Show that both formulations of $D_{i}$ are equivalent.


## Calculating PNL (6 of 8)

*Daily as Change in Cummulative:* **SELL**


\begin{align*}
C_{i} &= P - A_{i} \\[10pt]
D_{i} &= \begin{cases} C_{1} & i = 1 \\ C_{i} - C_{i-1} & i > 1 \end{cases}
\end{align*}

&nbsp;

**Exercise:** Show that both formulations of $D_{i}$ are equivalent.



## Calculating PNL (7 of 8)

- Suppose on 9/16/2013 we buy the 169 SPY expiring 9/21, paying EOD Ask.  

- Here is the price data for that option until expiration:
```{r, echo=FALSE}
library(tidyverse)
library(dplyr)
df_option_history <- read_csv("data_practice_option_history.csv")

df_put_hist <-
  df_option_history %>%
    dplyr::filter(underlying == "SPY") %>%
    dplyr::filter(expiration == "2013-09-21") %>%
    dplyr::filter(trade_date >= "2013-09-11") %>%
    dplyr::filter(type == "put") %>%
    dplyr::filter(strike == 169) %>%
    arrange(trade_date) %>%
    #select(trade_date, underlying, type, expiration, strike, upx, bid, ask)
    select(trade_date, bid, ask)
df_put_hist
```


## Calculating PNL (8 of 8)

**Exercise:** Verify the following PNL calculations.
```{r}
df_put_hist <-
  df_put_hist %>% 
    mutate(D_i = 0, C_i = 0)

for (ix_date in 1:nrow(df_put_hist)){
  if (ix_date == 1) {df_put_hist$D_i[ix_date] <- df_put_hist$bid[ix_date] - df_put_hist$ask[ix_date]}
  else {df_put_hist$D_i[ix_date] <- df_put_hist$bid[ix_date] - df_put_hist$bid[ix_date - 1]}
  
  df_put_hist$C_i[ix_date] = sum(df_put_hist$D_i[1:ix_date])
}
df_put_hist
```



## Black-Scholes-Merton Formula (1 of 1)
\vspace{-25pt}
\begin{align*}
m = \text{BSM}(\text{p/c}, K, \text{T}, S_{t},\sigma, \delta, r)
\end{align*}

**Contract Features**

- p/c - put or call

- $K$ - strike price

- T - expiration date (time to expiration)

**Market Values**

- $S_{t}$ - current underlying price

- $\sigma$ - estimate of the standard deviation log-return of the price of underlying between now and expiration

- $\delta$ - estimate of dividends paid over the life of the option

- $r$ - risk-free interest rate

## Black-Scholes-Merton Formula (2 of 2)

- For this class, we won't need to examine the BSM formula too closely, or write a BSM function from scratch.
\vspace{10pt}
- A more important skill wil be to use pre-existing BSM functions (and implied volatility functions) in analysis code.
\vspace{10pt}
- I usually use the `fOptions` package.
\vspace{10pt}
- **Exercise:**  Install the `fOptions` package and quickly read through pp 22-27 of its documentation PDF. 

## Greeks Preview (1 of 4)

```{r, echo=FALSE, fig.height = 4.5, fig.width=6.67, out.width="100%"}
library(fOptions)
library(bizdays)
library(tidyverse)
library(gridExtra)

type = "c"
upx = seq(90, 110, 1)
strike = 100
time = 21/252
rf = 0.01
div = 0
vol = 0.25

option_values <-
  GBSCharacteristics(
    TypeFlag = type
    , S = upx
    , X = strike
    , Time = time
    , r = rf
    , b = div
    , sigma = vol
  )


graph_values <-
  tibble(
    upx = upx
    , intrinsic = map_dbl(upx, function(x) max(x - strike, 0))
    , px = option_values$premium
    , delta = option_values$delta
    , vega = option_values$vega / 100
    , theta = option_values$theta / 252
  )


p1 <-
ggplot(data = graph_values ) +
  geom_line(aes(x = upx, y = intrinsic), size = 0.75) +
  geom_line(aes(x = upx, y = px), color = "green4") +
  geom_vline(xintercept = 100, color = "red", lty = "dashed") +
  labs(
       x = ""
       ,y = ""
  ) +
  annotate("text", label = "option price" , x = 93, y = 9.99, size = 4, colour = "black")

p2 <-
ggplot(data = graph_values ) +
  geom_line(aes(x = upx, y = delta), color = "blue") +
  geom_vline(xintercept = 100, color = "red", lty = "dashed") +
  labs(
       x = ""
       , y = ""
  ) +
  annotate("text", label = "delta" , x = 91.5, y = 0.875, size = 4, colour = "black")


p3 <-
ggplot(data = graph_values ) +
  geom_line(aes(x = upx, y = vega), color = "orange") +
  geom_vline(xintercept = 100, color = "red", lty = "dashed") +
  labs(
       x = "underlying price"
       , y = ""
  ) +
  annotate("text", label = "vega" , x = 91.5, y = .11, size = 4, colour = "black")

p4 <-
ggplot(data = graph_values ) +
  geom_line(aes(x = upx, y = theta), color = "pink") +
  geom_vline(xintercept = 100, color = "red", lty = "dashed") +
  labs(
       x = "underlying price"
       , y = ""
  ) +
  annotate("text", label = "theta" , x = 93, y = -0.025, size = 4, colour = "black")



gridExtra::grid.arrange(p1, p2, p3, p4
                        , nrow = 2
                        , top = "Greeks vs Underlying Price")
```


## Greeks Preview (2 of 4)
```{r, echo=FALSE}
library(fOptions)
library(tidyverse)
library(gridExtra)


type = "c"
upx = seq(90, 105, 1)
strike = 100
time = c(21, 10, 5, 1)
rf = 0.01
div = 0
vol = 0.25


df_graph_data <- tibble(
  desc = "0"
  , upx = upx
  , value = map_dbl(upx, function(x) max(x - strike, 0))
)


for (ix_time in 1: length(time)){
## for ix_time start ##########
option_values <-
  GBSCharacteristics(
  TypeFlag = type
  , S = upx
  , X = strike
  , Time = time[ix_time]/252
  , r = rf
  , b = div
  , sigma = vol
  )


df_add_values <- tibble(
  desc = as.character(time[ix_time])
  , upx = upx
  , value = option_values$premium
)

df_graph_data <-
  bind_rows(df_graph_data, df_add_values)
## for ix_time end ############
}

# this is the plot mothafucka
ggplot() +
  geom_line(
            data = df_graph_data %>% dplyr::filter(desc != "0")
            , aes(x = upx, y = value, color = desc)
  ) +
  scale_color_discrete(
                       name = "trade days\nto expiration"
                       , breaks = c("21", "10", "5", "1")
                       , labels = c("21", "10", "05", "01")
  ) +
  geom_line(
            data = df_graph_data %>% dplyr::filter(desc == "0")
            , aes(x = upx, y = value)
            , size = 0.75
  ) +
  theme(legend.position = c(0.15,0.65)) +
  labs(
       title = "The Effect of Time on Option Value"
       , x = "underlying price"
       , y = "option price"
  )
```



## Greeks Preview (3 of 4)
```{r, out.width="100%"}
library(fOptions)
library(bizdays)
library(tidyverse)
library(gridExtra)


type = "c"
upx = seq(90, 105, 1)
strike = 100
time = 21/252
rf = 0.01
div = 0
vol = seq(0.10, 0.30, 0.05)


df_graph_data <- tibble(
  vol = 0
  , upx = upx
  , value = map_dbl(upx, function(x) max(x - strike, 0)) 
)


for (ix_vol in 1: length(vol)){
## for ix_vol start ##########
option_values <- 
  GBSCharacteristics(
  TypeFlag = type
  , S = upx
  , X = strike
  , Time = time
  , r = rf
  , b = div
  , sigma = vol[ix_vol]
  )


df_add_values <- tibble(
  vol = vol[ix_vol]
  , upx = upx
  , value = option_values$premium
)

df_graph_data <-
  bind_rows(df_graph_data, df_add_values)
## for ix_time end ############
}

# this is the plot mothafucka
ggplot() +
  geom_line(
            data = df_graph_data %>% dplyr::filter(vol != 0)
            , aes(x = upx, y = value, color = factor(desc(vol)))
  ) +
  scale_color_discrete(
                       name = "volatility"
                       # , breaks = c("21", "10", "5", "1")
                       , labels = c("0.30", "0.25", "0.20", "0.15", "0.10")
  ) +
  geom_line(
            data = df_graph_data %>% dplyr::filter(vol == 0)
            , aes(x = upx, y = value)
            , size = 0.75
  ) +
  theme(legend.position = c(0.15,0.65)) +
  labs(
       title = "The Effect of Volatility on Option Value"
       , x = "underlying price"
       , y = "option price"
  )

```


## Greeks Preview (4 of 4)
**Some Facts**

1. Optionality - as evidenced by vega, theta, gamma - is greatest when options are ATM.
\vspace{10pt}
2. Theta is Negative: an option loses value as it nears expiration.
\vspace{10pt}
3. Vega is Positive: the more volatile the underlying, the more valuable the option.
\vspace{10pt}
4. Regarding Delta:
    - Approximiately 0.50 when option is ATM.  
    - Approaches 0.00 as option gets farther out of the money.  
    - Approaches 1.00 as option goes farther in the money.
    - *VERY roughly* the probability that the option expires ITM.
    - Used to refer to the moniness of an option.

